Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-01-08 article-models-eg.tex
moved some stuff around
working on proposal

\section{Empirical evidence of what an internal model should be}

@@

\begin{defn}
A \emph{graph $A$ in the context $\Gamma$} is a pair $\pairr{A_0,A_1}$ consisting
of 
\begin{align*}
A_0 & :\Gamma_0\to\type\\
A_1 & :\prd*{i,j:\Gamma_0}\Gamma_1(i,j)\to A_0(i)\to A_0(j)\to\type.
\end{align*}
Thus, for a graph $\Gamma$, the type $\mftyp{\mfGraph}{\Gamma}$ is the type
\begin{equation*}
\sm{A_0:\Gamma_0\to\type}\prd*{i,j:\Gamma_0}\Gamma_1(i,j)\to A_0(i)\to A_0(j)\to\type.
\end{equation*}
We will also write $\Gamma\vdash A:\mfGraph$ when $A$ is a graph in context
$\Gamma$.
\end{defn}

\begin{defn}
Suppose that $\Gamma\vdash A:\mfGraph$. Then we define the graph $\ctxext{\Gamma}{A}$
by
\begin{align*}
\ctxext{\Gamma}{A}_0 & \defeq \sm{i:\Gamma_0}A_0(i)\\
\ctxext{\Gamma}{A}_1(\pairr{i,x},\pairr{j,y}) & \defeq \sm{q:\Gamma_1(i,j)}A_1(q,x,y).
\end{align*}
\end{defn}

@@

@@

\subsubsection{The contexts of $\mfGraph$}

\begin{defn}
Suppose that $\Delta\defeq \pairr{\Delta_0,\Delta_1}$ and $\Gamma\defeq \pairr{\Gamma_0,
\Gamma_1}$ are graphs. A morphism $f$ from $\Delta$ to $\Gamma$ is a pair
$\pairr{f_0,f_1}$ consisting of
\begin{align*}
f_0 & : \Delta_0\to\Gamma_0\\
f_1 & : \prd*{u,v:\Delta_0}\Delta_1(u,v)\to\Gamma_1(f_0(u),f_0(v))
\end{align*}
We write $\ctxhom{\Delta}{\Gamma}$ for the type of morphisms of graphs.
\end{defn}

The terminal graph $\unit^\mfGraph$ plays the role of the empty context.

It is immediate that the type of graph morphisms $\Gamma\to\unit^\mfGraph$ is
contractible for every graph $\Gamma$. The terms of a graph $\Gamma$ are the
global elements of $\Gamma$, i.e.~they are the morphisms of type 
$\ctxhom{\unit^\mfGraph}{\Gamma}$.

\begin{defn}
Let $\Gamma$ be a graph. The type $\terms[\mfGraph]{\Gamma}$ of terms of $\Gamma$ is
defined to be
\begin{equation*}
\sm{x_0:\Gamma_0}\Gamma_1(x_0,x_0).
\end{equation*}
\end{defn}

\subsubsection{Families of graphs}
Using $\gobjclass$, we obtain a description of what
it means to be a graph in a context $\Gamma$, where $\Gamma$ is itself a
graph. The notion of ``a graph in a context $\Gamma$'' is the interpretation
of ``a type in a context $\Gamma$''.

\begin{defn}
A \emph{graph $A$ in the context $\Gamma$} is a pair $\pairr{A_0,A_1}$ consisting
of 
\begin{align*}
A_0 & :\Gamma_0\to\type\\
A_1 & :\prd*{i,j:\Gamma_0}\Gamma_1(i,j)\to A_0(i)\to A_0(j)\to\type.
\end{align*}
We will also write $\Gamma\vdash A:\mfGraph$ when $A$ is a graph in context
$\Gamma$.
\end{defn}

\begin{rmk}
Note that every graph can be seen as a graph in context $\unit^\mfGraph$.
\end{rmk}

\begin{defn}
Let $A$ be a graph in context $\Gamma$ and let $f:\ctxhom{\Delta}{\Gamma}$
be a graph morphism. We define the graph $A[f]$ in context $\Delta$ by
\begin{align*}
A[f]_0(u) & \defeq A_0(f_0(u))\\
A[f]_1(p,x,y) & \defeq A_1(f_1(p),x,y).
\end{align*}
\end{defn}

\begin{defn}
Let $\Gamma\vdash A:\mfGraph$ and $\Gamma\vdash B:\mfGraph$ be two graphs in
the context $\Gamma$. A morphism $f$ from $A$ to $B$ is a pair $\pairr{f_0,f_1}$
consisting of
\begin{align*}
f_0 & : \prd*{i:\Gamma_0} A_0(i)\to B_0(i)\\
f_1 & : \prd*{i,j:\Gamma_0}{q:\Gamma_0}*{x:A_0(i)}*{y:A_0{j}}A_1(q,x,y)\to B_1(q,f_0(x),f_0(y)).
\end{align*}
We write $\ctxhom[\Gamma]{A}{B}$ for the type of morphisms between graphs in
context $\Gamma$.
\end{defn}

\begin{defn}
The unit graph $\unit^\Gamma$ in context $\Gamma$ consists of
\begin{align*}
(\unit^\Gamma)_0 & \defeq \lam{i}\unit\\
(\unit^\Gamma)_1(q) & \defeq \lam{x}{y}\unit.
\end{align*}
\end{defn}

As was the case with graphs, terms of graphs in a context are described by
morphisms from $\unit$. Explicating this in more detail, we get

\begin{defn}
Let $A$ be a graph in context $\Gamma$. A term $x$ of $A$ is a pair
$\pairr{x_0,x_1}$ consisting of
\begin{align*}
x_0 & : \prd{i:\Gamma_0}A_0(i)\\
x_1 & : \prd*{i,j:\Gamma_0}{q:\Gamma_1(i,j)}A_1(q,x_0(i),x_0(j)).
\end{align*}
We define $\terms[\mfGraph]{A}$ to be the type of terms of $A$.
\end{defn}

\begin{rmk}
Note that the type of terms of a graph $\Gamma$ is equivalent to the type
of terms of the graph $\Gamma$ in context $\unit^\mfGraph$.
\end{rmk}

\begin{defn}
Suppose that $\Gamma\vdash A:\mfGraph$. Then we define the graph $\ctxext{\Gamma}{A}$
by
\begin{align*}
(\ctxext{\Gamma}{A})_0 & \defeq \sm{i:\Gamma_0}A_0(i)\\
(\ctxext{\Gamma}{A})_1(\pairr{i,x},\pairr{j,y}) & \defeq \sm{q:\Gamma_1(i,j)}A_1(q,x,y).
\end{align*}
There is a morphism $\pi^A:\ctxhom{\ctxext{\Gamma}{A},\Gamma}$ given by
\begin{align*}
\pi^A_0 & \defeq \lam{\pairr{i,x}}i && : (\sm{i:\Gamma_0}A_0(i))\to \Gamma_0\\
\pi^A_1(\pairr{i,x},\pairr{j,y}) & \defeq \lam{\pairr{q,e}}q && : (\sm{q:\Gamma_1(i,j)}A_1(q,x,y))\to\Gamma_1(i,j).
\end{align*}
\end{defn}

\begin{rmk}
The construction of $\ctxext{\Gamma}{A}$ is similar to the Grothendieck
construction of the category of elements for a presheaf.
\end{rmk}

\begin{defn}
Let $\Gamma\vdash A:\mfGraph$. A family $P$ of graphs over $A$ is by
definition a graph $P$ in the context $\ctxext{\Gamma}{A}$. More explicitly, a family
$P$ of graphs over $A$ in context $\Gamma$ consists of
\begin{align*}
P_0 & : \prd*{i:\Gamma_0}A_0(i)\to\type\\
P_1 & : \prd*{i,j:\Gamma_0}{q:\Gamma_1(i,j)}*{x:A_0(i)}*{y:A_0(j)} A_1(q,x,y)\to P_0(x)\to P_0(y)\to\type
%P_1 & : \prd{\pairr{i,x},\pairr{j,y}:(\ctxext{\Gamma}{A})_0}(\ctxext{\Gamma}{A})_1(\pairr{i,x},\pairr{j,y})\to P_0(i,x)\to P_0(j,y)\to\type.
\end{align*}
\end{defn}

\begin{defn}
When $P$ is a family of graphs over $A$ in context $\Gamma$ and $x:A$, we
get a graph $\subst{x}{P}$ in context $\Gamma$ given by
\begin{align*}
\subst{x}{P}_0(i) & \defeq P_0(x_0(i))\\
\subst{x}{P}_1(q,u,v) & \defeq P_1(\pairr{q,x_1(q)},u,v)
\end{align*}
\end{defn}

\subsubsection{The basic type constructors for graphs}
With families of graphs being available, we can give the interpretations of
dependent products, dependent sums and identity types. In the following, we
shall introduce the graph interpretations of dependent products, dependent
sums and identity types and describe the terms of the resulting graphs.\note{These definitions
should be connected to Mike's article but I don't really know how to do this}

@@

\begin{rmk}
For any two graphs $A$ and $B$ in a context $\Gamma$, we have the graph
$B[\pi^A]$ in the context $\ctxext{\Gamma}{A}$, which we may just write
as $B$. Therefore, we can
consider the graphs $\mprd{A}{B}$ and $\msm{A}{B}$. As usually those shall be
denoted  by $A\to B$ and $A\times B$, respectively.

A term $f$ of $A\to B$ consists of
\begin{align*}
f_0 & : \prd*{i:\Gamma_0}A_0(i)\to B_0(i)\\
f_1 & : \prd*{i,j:\Gamma_0}{q:\Gamma_1(i,j)}*{x:A_0(i)}*{y:A_0(j)}A_1(q,x,y)\to B_1(q,f_0(x),f_0(y)),
\end{align*}
so we see that $\terms{A\to B}\jdeq\ctxhom[\Gamma]{A}{B}\jdeq\terms{B[\pi^A]}$.

A term $\pairr{x,y}$ of $A\times B$ consists of
\begin{align*}
\pairr{x,y}_0 & : \prd*{i:\Gamma_0} A_0(i)\times B_0(i)\\
\pairr{x,y}_1 & : \prd*{i,j:\Gamma_0}{q:\Gamma_1(i,j)} A_1(q,x_0(i),x_0(j))\times B_1(q,y_0(i),y_0(j))
\end{align*}
so we see that $\eqv{\terms{A\times B}}{\terms{A}\times\terms{B}}$.

Note that we also obtain
a \emph{graph} $\Delta\to\Gamma$ this way, for every two graphs $\Delta$ and
$\Gamma$ seen as graphs in the context $\unit^\mfGraph$. 
The type $\terms{\Delta\to\Gamma}$ of terms of $\Delta\to\Gamma$ is just
$\ctxhom{\Delta}{\Gamma}$.
\end{rmk}

\begin{defn}
There is a graph morphism
\begin{equation*}
\graphcharmapfunc{\Gamma} : \sm{\gobjclass}
\end{equation*}
\end{defn}

@@

\subsubsection{Properties of graphs}

>>>>

\section{Empirical evidence of what an internal model should be}\label{eg}

@@

\begin{defn}
A \emph{family $A$ of graphs in the context $\Gamma$} is a pair $\pairr{A_0,A_1}$ consisting
of 
\begin{align*}
A_0 & :\Gamma_0\to\type\\
A_1 & :\prd*{i,j:\Gamma_0}\Gamma_1(i,j)\to A_0(i)\to A_0(j)\to\type.
\end{align*}
Thus, for a graph $\Gamma$, the type $\mftyp{\mfGraph}{\Gamma}$ is the type
\begin{equation*}
\sm{A_0:\Gamma_0\to\type}\prd*{i,j:\Gamma_0}\Gamma_1(i,j)\to A_0(i)\to A_0(j)\to\type.
\end{equation*}
We will also write $\Gamma\vdash A:\mfGraph$ when $A$ is a graph in context
$\Gamma$.
\end{defn}

\subsubsection{The interpretations of extension, weakening and substitution}
\begin{defn}
Suppose that $\Gamma\vdash A:\mfGraph$. Then we define the graph $\ctxext{\Gamma}{A}$
by
\begin{align*}
\ctxext{\Gamma}{A}_0 & \defeq \sm{i:\Gamma_0}A_0(i)\\
\ctxext{\Gamma}{A}_1(\pairr{i,x},\pairr{j,y}) & \defeq \sm{q:\Gamma_1(i,j)}A_1(q,x,y).
\end{align*}
\end{defn}

@@

\begin{rmk}
Using weakening we can describe the graph morphisms. Suppose that $\Delta$ and
$\Gamma$ are graphs. A graph morphism from $\Gamma$ to $\Delta$ is a term of 
the family $\ctxwk{\Gamma}{\Delta}$ of graphs over $\Gamma$. More explicitly,
a graph morphism $f$ is a pair
\begin{align*}
\pts{f} & : \pts{\Gamma}\pts{\Delta}\\
\edg{f} & : \prd*{i,j:\pts{\Gamma}} \edg{\Gamma}(i,j)\to\edg{\Delta}(\pts{f}(i),\pts{f}(j))
\end{align*}
just as expected.

We may also describe morphisms of families of graphs this way.
Let $A$ and $B$ be two families of graphs over a graph $\Gamma$. 
A morphism from $A$ to $B$ is a term of $\ctxwk{A}{B}$. More explicitly,
a morphism $f$ from $A$ to $B$ is a pair $\pairr{f_0,f_1}$
consisting of
\begin{align*}
\pts{f} & : \prd*{i:\Gamma_0} A_0(i)\to B_0(i)\\
\edg{f} & : \prd{\pairr{i,x},\pairr{j,y}:\pts{\ctxext{\Gamma}{A}}}{\pairr{q,e}:\edg{\ctxext{\Gamma}{A}}(\pairr{i,x},\pairr{j,y})} \edg{B}(q,f_0(x),f_0(y)).
\end{align*}
Here we have suppressed the notation for the projections.
\end{rmk}

\subsubsection{The interpretations of the type constructors}

@@

With families of graphs being available, we can give the interpretations of
dependent products, dependent sums and identity types. In the following, we
shall introduce the graph interpretations of dependent products, dependent
sums and identity types and describe the terms of the resulting graphs.\note{These definitions
should be connected to Mike's article but I don't really know how to do this}

@@

@@

\begin{comment}
\subsubsection{The contexts of $\mfGraph$}

\subsubsection{The basic type constructors for graphs}

\begin{defn}
There is a graph morphism
\begin{equation*}
\graphcharmapfunc{\Gamma} : \sm{\gobjclass}
\end{equation*}
\end{defn}
\end{comment}

### 2014-01-08 article-models-intro.tex
working on proposal

\section{Introduction}
The project I propose here has its origins in the beginning of 2013, when I proved a version
of the descent theorem for homotopy colimits in type theory while I was working
with Bas Spitters to develop notions from higher category theory in the
univalent foundations. To arrive at a
notion of diagram general enough to capture the higher inductive types described
in chapter 6 of \cite{TheBook} excluding the truncations we needed type
theoretical graphs. The graphs form a model of type theory and indeed we needed
several of the basic type theoretical operations to give an efficient approach
to the descent theorem. Although it was not an issue to describe the graph model
and the sense in which it models the type constructors, 
not all of type theory is interpreted very well. To start with, context
extension isn't strictly associative if dependent pair types are used for
its interpretation. Also, if $A$ and $B$ are families of graphs
over a graph $\Gamma$, then $B$ isn't also a family of graphs over the extended
graph $\ctxext{\Gamma}{A}$ (interfering with a good interpretation of weakening).
There are no issues with the interpretation of substitution though. Actually
it is not so straightforward to describe what it models!

The graph model
shows three operations: extension, weakening and substitution. All of them
can be defined such that they act not only on contexts, but also on families
and terms. Moreover, they are all compatible with each other. In fact, along
with asserting the existence of identity morphisms (which is done in type
theory), it seems
that this structure is enough to establish some of the structure of a category.
We might even go so far as to boldly assert that
\begin{quote}
\emph{Category theory is dependent type theory without type constructors.}
\end{quote}
It is a part of this proposal to test this hypothesis, with the note that by
`category theory' we mean a theory of weak higher categories which does not
have any of the truncatedness restrictions -- in particular not those appearing
in the theory of AKS-categories as
described in \cite{TheBook} in chapter 11. We note that even Dybjers theory
of internal categories with families has such restrictions: he relies on setoids.
It needs not much arguing that these are very unnatural from a univalent point
of view.

\subsection{Ideas in the definition of internal models}
An internal model of type theory is like a category with families, but we want
to avoid having to state higher coherences. In fact, we don't even start our
definition with a category of contexts; instead we just take a \emph{type} of contexts. 
The morphisms will come from the terms, evaluation of a function at a given
term will come from substitution. We recognize three basic ingredients to models:
first there is a type of contexts; second, for every context there is a model of types in
that context and third, for every type in a given context there is a type of its
terms. Then there are three basic attributes: context extension, weakening and
substitution. Context extension provides us with families over types as well as
with an interpretation of dependent pair types. We need weakening 
so that families can depend on the same type multiple times (the way the
identity type of a type depends two times on that type) and to be able
to talk about non-dependent function types,
the morphims of our category. Substitution will give us a way
to work with fibers of families as well as composition of functions and evaluation
of functions at terms.

@@

When we have this framework set up, we can interpret the basic type constructors
such as $\Pi$, $\Sigma$ and $\idtypevar{}$.
The higher categorical structure then comes from the
result that we have an interpretation of type theory.

\subsection{The elementary theory of the category with families of categories
with families}

Thus the idea was born to give a new description of Martin-L\"of type theory
which is more faithful to the idea that type theory is an algebraic description
of higher category theory. The features we have in mind for this description
is that:
\begin{enumerate}
\item As in Martin-L\"of type theory there are the three basic judgments
      \begin{align*}
        & \Gamma\text{ is a context:} & \jctx*{\Gamma}\\
        & A\text{ is a type in context }\Gamma\text{:} & \jtype*{\Gamma}{A}\\
        & x\text{ is a term of $A$ in context }\Gamma\text{:} & \jterm*{\Gamma}{A}{x}
      \end{align*}
      together with the three accompanying judgments for judgmental equality.
      Our new description of type theory should be such that we can faithfully
      make the translation to the three assertions
      \begin{enumerate}
      \item $\Gamma$ is an object\\
      \item $A$ is a fibration over $\Gamma$\\
      \item $x$ is a section of the fibration $A$. 
      \end{enumerate}
\item There is a basic type theory which has just the operations
      \begin{description}
      \item[extension] taking the domain of the fibration $A$ over $\Gamma$.
      \item[weakening] given an object $\Gamma$, weakening assigns to an object
      $\Delta$ trivial fibration with $\Delta$ as its fibers.
      \item[substitution] taking the fiber at a point. Fibers should always
      be equivalent to homotopy fibers.
      \end{description}
      Thus we aim for a closer connection between the syntactic operations and
      the operations of models.
\item Models of this basic type theory should be weak $\omega$-categories. Thus,
      it should not be necessary to start the definition of a model with a bunch
      of contexts and substitution morphisms between them: these and the
      operations on them, such as composition and associativity thereof, should
      be a consequence of the interpretation of the basic type theory.
\item There shouldn't be an empty context and dependent function types right
      from the start. These correspond to having a terminal object and the
      models of type theory with dependent function types are the locally
      cartesian closed models.
\item Contexts should not be lists of variable declarations and all the
      operations should be explicitly invariant under context extension.
\item There should be a large zoo of extensions of the basic type theory. First
      off, we can extend the basic type theory with identity types, dependent
      function types and inductive types. We want a general description of
      homotopy colimits in type theory, which should become available once we
      know what categories (models of the basic type theory) are. Other types
      such as the truncations could be added as well.

      An extension of a different kind is the addition of the univalence axiom.
\end{enumerate}

\subsection{The further goals of the project}
The first aim of this project is thus to give this new description of type
theory with these properties. The second aim is to define a general notion of
an internal model and provide several examples thereof. In the definition of
internal models we envision that all the operations of type theory become
morphisms of type theory: they are compatible with each and every morphism
of type theory. The compatibility of extension, weakening and substitution is
already part of the basic type theory. The compatibility of identity types
with weakening for instance, is a form of function extensionality. The 
compatibility of dependent function types with identity types is the usual
funchtion extensionality principle. The
compatibility of a universe operation with identity types should be univalence.
The descent theorem implies that the colimit operation is also compatible with
all the type theoretic operations (sigma and id are checked).

Among the examples of internal models we should have:
\begin{description}
\item[The setoid model] This is a classical one so it should be there. It might
      be a bit different in our case. We'll want to construct a setoid model
      of the basic type theory to interpret identity types without necessarily
      interpreting dependent function types (but that should remain possible).
\item[The graph model] This is the model that leads to the first version of the
      descent theorem.
\item[Univalent unvirses] A univalent universe should be a model.
\item[Equifibered diagrams over a graph] should be the fibrations in some model.
\item[The model of all models] Very likely, we will even define this simultaneously
      with the notion of model.
\item[The model of weak $\omega$-groupoids] Since we think of models of the
      basic type theory as weak $\omega$-categories, it is not too hard to
      provide a condition on those models which enables us to talk about
      weak $\omega$-categories. In fact, we have several options for such a
      condition. We investigate those; we also investigate how they relate to
      Brunerie's weak $\omega$-groupoids. Presumably, they form an internal
      model without too much fuss (because we already have an internal model
      of internal models at this stage) and we can ask whether this models
      the univalence axiom. In fact, the model of internal models might already
      have modeled the univalence axiom. 
\end{description}

@@

\subsection{Overview of the document}
In \autoref{eg} we begin the project with exploring internal models of type theory
in type theoretical setting of \cite{TheBook}.

>>>>

\section{Stage one: establishing internal higher categories and internal models}
The project I propose here has its origins in the beginning of 2013, when I proved a version
of the descent theorem for homotopy colimits in type theory while I was working
with Bas Spitters to develop notions from higher category theory in the
univalent foundations. To arrive at a
notion of diagram general enough to capture all the higher inductive types described
in chapter 6 of \cite{TheBook} excluding the truncations we needed type
theoretical graphs. The graphs form a model of type theory and indeed we needed
the interpretations of several of the basic type constructors to give an efficient approach
to the descent property and its proof. Although it was not an issue to describe the graph model
and the sense in which it models the type constructors, 
not all of type theory is interpreted very well. To start with, if $A$ and $B$ 
are families of graphs over a graph $\Gamma$, then $B$ isn't also a family of 
graphs over the extended graph $\ctxext{\Gamma}{A}$ (interfering with a good 
interpretation of weakening). Since there are such issues, it is actually
not so straightforward to describe what it models!

Apart from all the type constructors like dependent function and pair types and
identity types, the graph model shows three basic operations that should come
before those type constructors. They are extension, weakening and substitution.
All three of them can be defined such that they act not only on contexts (the graphs), 
but also on families and terms. Moreover, they are all compatible with each 
other. In fact, any structure which has a notion of contexts, families and terms
and which models extension, weakening and substitution and which has identity
functions is in a weak sense a category, hypothesizing that
\begin{quote}
\emph{Category theory is dependent type theory without type constructors.}
\end{quote}
In \autoref{tt} we sketch how type theory without type constructors but with
\emph{explicit} extension, weakening and substitution might look like. This
theory is written down keeping the guiding principle in mind that type theory
without basic constructors should be a generalized algebraic theory of higher
categories.

Thus, models of this theory should be higher categories with some strictness
built in, such as the strictness of associativity of substitution. But the
theory really only talks about contexts, families and sections. The interpretation
of this as a category works as follows: The objects are going to be the contexts.
Then we may consider the terms of the weakening $\ctxwk{\Gamma}{\Delta}$ of a
context $\Delta$ by a context $\Gamma$: they are the morphisms of the category.
The composition of morphisms follows from the action of substitution and
weakening on terms.

In the proposed approach, a model of type theory isn't a category by assumption. Rather,
a model of type theory \emph{becomes} a category because it interprets type
theory. In other words, type theory does all the work for us concerning the
categorical structure. In particular, we have neither started our notion of models
with AKS-categories as described in \cite{1categories} nor with Dybjer's notion
of internal categories with families. We should gain that we get rid of
truncatedness assumptions that are present in both AKS-categories, where the
type of objects has to be a $1$-type, and internal categories with families,
where the type of morphisms has to be a setoid. It needs not much arguing that any way to impose
conditions on internal models beyond the interpretations of the rules of
type theory stands in the way of a clear understanding of what an internal
model should be and it is a main goal of the project to design
an internal type theory without the described deficiencies.

\begin{comment}
\subsection{Ideas in the definition of internal models}
An internal model of type theory is like a category with families, but we want
to avoid having to state higher coherences. In fact, we don't even start our
definition with a category of contexts; instead we just take a \emph{type} of contexts. 
The morphisms will come from the terms, evaluation of a function at a given
term will come from substitution. We recognize three basic ingredients to models:
first there is a type of contexts; second, for every context there is a model of types in
that context and third, for every type in a given context there is a type of its
terms. Then there are three basic attributes: context extension, weakening and
substitution. Context extension provides us with families over types as well as
with an interpretation of dependent pair types. We need weakening 
so that families can depend on the same type multiple times (the way the
identity type of a type depends two times on that type) and to be able
to talk about non-dependent function types,
the morphims of our category. Substitution will give us a way
to work with fibers of families as well as composition of functions and evaluation
of functions at terms.

@@

When we have this framework set up, we can interpret the basic type constructors
such as $\Pi$, $\Sigma$ and $\idtypevar{}$.
The higher categorical structure then comes from the
result that we have an interpretation of type theory.
\end{comment}

Of course there is no way of studying internal models or categories without
also studying the morphisms between them. In fact, we already have three of them
right from the start: the interpretations of extension, weakening and substitution
act on contexts, families and terms and preserve each of extension, weakening and
substitution. These have to be ingredients of morphisms of models as well. A
morphism of models should act on contexts because those are the contexts; it
should act on terms because that is how it acts on the morphisms; it should
preserve extension, weakening and substitution because that is exactly what
makes it functorial (in particular: preserve composition).

In fact, now that we have arrived at the idea that categorical structure should
come from the interpretation of type theory, the prettier (and necessary) way to go is to
describe a model of models of type theory. Thus, we should find a way to express
the notions of dependent model and terms thereof and explain what extension,
weakening and substitution mean. Then we should be able to derive that a morphism
of models coincides with a term of a weakened object.

The idea that morphisms of models preserve the basic type theoretical structure
can be carried further. Ideally, we would have the interpretations of the type
constructors as morphisms of models. Thus, when interpreting the operations 
$\Pi$, $\Sigma$, $\idtypevar{}$ or the universe operator 
(such as described in \cite{Palmgren1998}) we will require that
they act not only on objects, but also on families, terms and preserve extension,
weakening and substitution. And when two or more type constructors are interpreted,
they should furthermore be compatible with each other. In the case of
dependent pair types and identity types, this means function extensionality. It
should be investigated whether the compatibility of identity types and the
universe operator with each other indeed means univalence. We should note,
however, that we intend the compatibility with extension, weakening and substitution
should be strict, whereas the compatibility with the other type constructors will
not be strict. To give an exact meaning to these ideas is part of the proposal.

Among the examples of internal models we should have:
\begin{description}
\item[The setoid model] This is a classical one so it should be there. It might
      be a bit different in our case. We'll want to construct a setoid model
      of the basic type theory to interpret identity types without necessarily
      interpreting dependent function types (but that should remain possible).
\item[The graph model] This is the model that leads to the first version of the
      descent theorem.
\item[Univalent unvirses] A univalent universe should be a model.
\item[Equifibered diagrams over a graph] should be the fibrations in some model.
\item[The model of all models] Very likely, we will even define this simultaneously
      with the notion of model.
\item[The model of weak $\omega$-groupoids] Since we think of models of the
      basic type theory as weak $\omega$-categories, it is not too hard to
      provide a condition on those models which enables us to talk about
      weak $\omega$-categories. In fact, we have several options for such a
      condition. We investigate those; we also investigate how they relate to
      Brunerie's weak $\omega$-groupoids. Presumably, they form an internal
      model without too much fuss (because we already have an internal model
      of internal models at this stage) and we can ask whether this models
      the univalence axiom. In fact, the model of internal models might already
      have modeled the univalence axiom. 
\item[The set model] See \cite{RijkeSpitters:Sets}.
\item[Polynomial functors]
\end{description}

\section{Stage two: extending the basic theory}
The first aim of this project is thus to give this new description of type
theory with these properties. The second aim is to define a general notion of
an internal model and provide several examples thereof. In the definition of
internal models we envision that all the operations of type theory become
morphisms of type theory: they are compatible with each and every morphism
of type theory. The compatibility of extension, weakening and substitution is
already part of the basic type theory. The compatibility of identity types
with weakening for instance, is a form of function extensionality. The 
compatibility of dependent function types with identity types is the usual
funchtion extensionality principle. The
compatibility of a universe operation with identity types should be univalence.
The descent theorem implies that the colimit operation is also compatible with
all the type theoretic operations (sigma and id are checked).

@@

### 2014-01-08 article-models-main.tex
moved some stuff around
Moved universal property of extension away since it is inclear whether we want it
working on the introduction
working on proposal

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Internal models of type theory}
\author{Egbert Rijke}
\date{\today}

@@

@@

\input{article-models-eg}

@@

\bibliographystyle{hplain}
%\phantomsection\addcontentsline{toc}{section}{References}
\bibliography{refs}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Internal models of type theory and internal higher categories for the
Univalent Foundations}
\author{Egbert Rijke}
\date{\today}

@@

\begin{abstract}
A project is proposed where we investigate notions of internal models of type
theory and internal higher categories. We take the point of view that an
internal model is an internal higher category with extra structure interpreting
the type constructors. We propose to (1) explore possible definitions of both
notions, (2) find a zoo of examples and (3) find extensions of the underlying
theory and models thereof to include higher inductive types.
\end{abstract}

@@

%\input{article-models-eg}

\appendix
\input{article-models-tt}

\input{article-models-idtypes}

%\input{article-models-constructors}

@@

\bibliographystyle{plain}
%\phantomsection\addcontentsline{toc}{section}{References}
\bibliography{refs}

### 2014-01-08 article-models-tt.tex
working on proposal

\section{Variable-free type theory -- the basic theory}
We have seen in the example of $\UU$ that we will not need to refer to variables
when we are manipulating contexts. In fact, it is more natural not to.
This suggests that there is a presentation
of type theory where contexts do not mention variables. The first small gain is that such a presentation
would not be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.

>>>>

\section{Type theory without type constructors}\label{tt}
We have seen in the example of $\UU$ that we will not need to refer to variables
when we are manipulating contexts. In fact, it is more natural not to.
This suggests that there is a presentation
of type theory where contexts do not mention variables. The first small gain is that such a presentation
would not be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.
